On The Subspace Learning for Network Attack Detection

### 2019-05-27 docs/thesis/chapters/4_m_rpca/m_rpca.tex
first deliverable

\chapter{Moment Distances from Robust Subspace for Botnet Detection}
\label{ch:4_m_rpca}

@@

Anomaly detection can be defined as the identification of rare and suspicions events by differing the normal or majority of the data. Anomalies are also referred to as outliers, novelties, noise or deviations, and can be related to network attacks, frauds or defects \cite{bhuyan2014network,ahmed2016survey}. Anomalies can be hard to identify and separate from normal data due to the rare occurrences of anomalies in comparison to normal events, therefore anomaly detection algorithms have to be high discriminative, robust to contamination and able to deal with the imbalanced data problem \cite{he2008learning}.

Several efforts and researches aim to avoid network attacks based on known attackers, fingerprints or other patterns. However, distributed attacks organized by botnet has increasing and demanding the development of counter measures in order to detect and avoid unknown attacks or even to deal with adversarial changes on behavior, location and other patterns, in order to avoid detection by behavioral or pattern based detection systems \cite{gu2008botminer, garcia2014empirical,khattak2015botflex,acarali2016survey,wang2017botnet,Wang2018ddosbotnetssurvey}. To face this problem, it is possible to adopt unsupervised or semi-supervised approaches for network anomaly detection, where it is not necessary known anomalies for training models \cite{moustafa2019holistic}.

Network anomaly detection problems are usually characterized by skewed and imbalanced data \cite{Phua2004minority,he2008learning,benson2010network}. Learning algorithms for imbalanced data has been a challenging research topic, considering that the fundamental issue with the imbalanced learning problem is the ability of imbalanced data to significantly compromise the performance of most standard learning algorithms \cite{he2008learning}. Some techniques can make the data discriminative for better results on the imbalanced learning problem: Data engineering (i.e. sampling \cite{he2008learning,gu2008botminer}, aggregation \cite{lakhina2005mining,gu2008botminer,callegari2011novel, garcia2014empirical, acarali2016survey}); Feature extraction (i.e. decomposition \cite{vaswani2018robust}, robust estimates \cite{zhou2017anomaly}).

% He and Garcia \cite{he2008learning} present a classification of imbalanced learning methods as sampling, cost-sensitive learning, kernel-based, and active learning. For more information about learning methods from imbalanced data, we refer to He and Garcia \cite{he2008learning}.

% Say that the data is lognormal \cite{benson2010network} and link it to skewness https://www.weibull.com/hotwire/issue47/relbasics47.htm
Some widely adopted algorithms assume a gaussian distributed data, however this assumption may not be observed in real world problems, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{lakhina2005mining,benson2010network}. Findings of Benson \emph{et al.}  \cite{benson2010network} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. This findings show that the skeweness and heavy-tailed distributions can impact algorithms that rely on gaussian distributed data, as well as it can reveal characteristics that can be exploited in order to obtain accurate classifiers for network anomaly detection. Therefore, learning methods for imbalanced and skewed data have been attracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}. Skewness and kurtosis are higher moments that can have their measurement as very sensitive for outlier detection, considering that the properties of the kurtosis has been used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection \cite{hubert2009robustskewed}.

Principal Component Analysis (PCA) is very sensitive to anomalies, thus outliers can corrupt the entire data or can be explored to highlight unexpected changes and indicate attacks or frauds \cite{callegari2011novel,Lee2013,vieira2017model}. However, PCA is not precise to reveal flows, time or detailed information for each anomaly, and sometimes relies into visual analysis for principal component selection. Robust Principal Component Analysis (RPCA) \cite{candes2011robust} is an extension of PCA that aims to be resilient to outliers by means of a robust subspace learning \cite{vaswani2018robust} for outlier corrupted data, decomposing a given data matrix $\textbf{X}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, which refers to outliers’ matrix. Therefore, robust subspace learning has receiving a growing attention of researchers aiming the development of network anomaly detection systems \cite{rousseeuw1984mcd, rousseeuw1999fastmcd, hubert2005robpca,hubert2009robustskewed, pascoal2012robust, zhou2017anomaly}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}.

We believe that the skewness of anomalous and normal traffic can highlight features for improving anomaly detection in imbalanced data. The botnet detection algorithms can also be improved by feature engineering from aggregated traffic, considering that new features can make the data more discriminative for imbalanced learning problems. We also believe that the distance between robust estimates of normal traffic and anomalies can highlight discrepancies and be used for botnet detection. Therefore, we propose an approach based on higher moments distances from robust subspace for botnet detection, applied to new features generated by flow aggregation. The proposed approach relies on a robust subspace of normal traffic for estimating higher moments. The anomaly detection for new observations evaluate the Mahalanobis distance between the robust higher moments and the higher moments of new observations, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations.

We propose a sampling method based on data aggregation for discriminative feature engineering, considering that data aggregation is a widely adopted approach for feature engineering \cite{garcia2014empirical, chandrashekar2014survey,acarali2016survey} and network anomaly detection \cite{lakhina2005mining, callegari2011novel, vieira2017model}. We evaluate the accuracy of the proposed approach for botnet detection on the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}.

Experimental evaluation compares our proposal to widely adopted algorithms for anomaly detection based on clustering and statistical approaches, which are K-means and Gaussian Mixture Model (GMM) \cite{gaddam2007kmeans,moustafa2019holistic}, respectively. We also evaluate the results of ROBPCA \cite{hubert2005robpca}, which is a method that also relies on robust estimates with adjusted outlyingness based on robust skewness.

The main contribution of this work is a novel semi-supervised method for botnet detection that obtain results with high rates of attack detection in an imbalanced and skewed data set.

This paper is organized as follows. In Section \ref{sec:review}, it is conducted a literature review of network anomaly detection, botnet detection, sampling and aggregation for imbalanced learning and robust based anomaly detection methods. Section \ref{sec:datamodel} presents the data model and the evaluated data set. Section \ref{sec:mdrs} describes the proposed approach for botnet detection. Section \ref{sec:experimentalresults} discusses the experimental validation and presents the results, and Section \ref{sec:conclusionandfutureworks} draws the conclusions and the suggestions for future work.

@@

% network anomaly detectiom
Bhuyan \emph{et al.} \cite{bhuyan2014network} provide an overview of facets of network anomaly detection, present attacks normally encountered by network intrusion detection systems, and categorize existing network anomaly detection methods and systems based on the underlying techniques. Ahmed \emph{et al.} \cite{ahmed2016survey} present an analysis of four major categories of anomaly detection techniques which include classification, statistical, information theory and clustering. Moustafa \emph{et al.} \cite{moustafa2019holistic} discuss aspects of anomaly-based Network Intrusion Detection Systems (NIDSs), explains cyber-attacks and new solutions for anomaly detection, and provides a benchmark data sets for training and validating approaches for detection.

@@

The objective of Garcia \emph{et al.} \cite{garcia2014empirical} is to compare three botnet detection methods using a simple and reproducible methodology, and a good data set and a new error metric. This paper evaluates some data sets for network anomaly detection, surveys some approaches for botnet detection, proposes two methods (BClus and CAMNEP) for botnet detection and compare results to BotHunter \cite{gu2007bothunter}. Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by attack, normal and background labeled data, in an imbalanced distribution like in a real network. The authors recommend that none of the botnet families used in the training and cross-validation data set should be used in the testing data set, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Adopting the training and testing approach proposed by Garcia \emph{et al.}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

@@

% Anomaly detection by robust subspace learning
The problem of PCA or subspace learning for outlier corrupted data is called robust PCA (RPCA) or robust subspace learning \cite{vaswani2018robust}, which can be defined as the decomposition of a given data matrix $\textbf{X}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, with outliers or noise. This definition is also referred to as the sparse + low rank (S + LR) formulation \cite{vaswani2018robust}. We refer to \cite{lerman2018overview} and \cite{vaswani2018robust} for more details regarding robust subspace learning.

Robust subspace learning has receiving a growing attention of researchers aiming the development of anomaly detection systems \cite{rousseeuw1984mcd, rousseeuw1999fastmcd}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}. RPCA is been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adopted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of mean and covariance matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness.

Pascoal \emph{et al.} \cite{pascoal2012robust} proposed an approach based on a robust mutual information estimator for feature selection and based on RPCA for outlier detection in internet traffic. The anomaly detection proposed by Pascoal \emph{et al.} is an unsupervised approach that estimate the first $k$ robust principal components, calculate the score and the orthogonal distances, calculate the thresholds and classify new observations accordingly. 	Zhou and Paffenroth \cite{zhou2017anomaly} proposed the Robust Deep Autoencoders (RDA), which the central idea is that a RDA inherits the non-linear representation capabilities of autoencoders combined with the anomaly detection capabilities of RPCA. Considering that outliers and noise may reduce the quality of representations discovered by deep autoencoders, the proposed model isolates noise and outliers in the input by means of a RPCA approach, and the autoencoder is trained after this isolation. RDA was evaluated by the authors for the MNIST data set.

% feature engineering and aggregation
Lakhina \emph{et al.} \cite{lakhina2005mining} argue that the distributions of packet features (IP addresses and ports) observed in flow traces reveals both the presence and the structure of a wide range of anomalies. Lakhina \emph{et al.} show that the analysis of feature distributions leads to significant advances on sensitive detection of a wide range of anomalies, augmenting detection by volume-based methods, while the anomaly detection approach leads gains on automatic classification of anomalies via unsupervised learning. Goldberg and Shan \cite{goldberg2015importance} argue, based on experience and on published real-life examples of anomaly detectors in use at eBay, that processing monitored signals into features is a more fruitful for anomaly detection systems than devising complex statistical tests.

Benson \emph{et al.} \cite{benson2010network} conducted an empirical study of the network traffic in 10 data centers belonging to three different types of organizations, including university, enterprise, and cloud data centers. Findings of Benson \emph{et al.} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. This findings highlight that the skeweness and heavy-tailed distributions can impact algorithms that rely on gaussian distributed data, as well as it can reveal characteristics that can be exploited in order to obtain accurate classifiers for network traffic analysis.

Considering that learning methods for imbalanced and skewed data have been attracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}, higher moments can be used for outlier detection and discriminative analysis of skewed data. The skewness and kurtosis are higher moments that can contribute to sensitive outlier detectors, taking into account that the properties of the kurtosis can be used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection in skewed data sets \cite{hubert2009robustskewed}.

We propose an approach based on higher moments distances from robust subspace for botnet detection, applied to new features generated by flow aggregation from the CTU-13 data set. The proposed approach relies on a robust subspace of normal traffic for estimating higher moments. Thus, we propose that the anomaly detection for new observations should evaluate the Mahalanobis distance between the robust higher moments and the higher moments of new observations, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations.

We propose a sampling method based on data aggregation for discriminative feature engineering, considering that data aggregation is a widely adopted approach for feature engineering \cite{garcia2014empirical, chandrashekar2014survey,acarali2016survey} and network anomaly detection \cite{lakhina2005mining, callegari2011novel, vieira2017model}. Finally, We evaluate the F-measure of the proposed approach for botnet detection on the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}.

@@

This section also presents a description of the data model in Subsection \ref{sec:CTU-13} and in Subsection \ref{sec:feat_eng} we describe the data aggregation and feature engineering adopted for the CTU-13.

\subsection{The CTU-13 data set}
\label{sec:CTU-13}

The CTU-13 is a data set of botnet traffic that was captured in the Czech Technical University \cite{garcia2014empirical}, by means of a testbed and execution of malwares in a real network. The CTU-13 data set contains 13 scenarios with network traffic of botnet malwares, that are: neris, rbot, virut, menti, sogou, nsys.ay and murlo. The types of C\&C and attack traffic present in CTU-13 data set are:

@@

The data set originally contains the following features by flows:  Start Time, End Time, Duration, Source IP Address, Source Port, Direction, Destination IP Address, Destination Port, State, Type of service from source to destination, Type of service from destination to source, Total of Packets and Total of Bytes.

Table \ref{tab:tab01} presents an overview organized by scenario and shows the malwares used for botnet attacks, the types of attacks and C\&C types, the total number of flows, the number of malicious flows, which includes flows of C\&C and attacks, and finally shows the number of normal flows.

\begin{table}[h!]
	\tiny
	\caption{CTU-13 data set Description}
	\label{tab:tab01}
	\begin{tabular}{| l | l | l | r | r | r | r | r | r | r | r | }
		\hline \rowcolor{Gray} \begin{tabular}[x]{@{}l@{}}\textbf{ID}\end{tabular}	&\begin{tabular}[x]{@{}l@{}}\textbf{Bot}\end{tabular}	 &\begin{tabular}[x]{@{}l@{}}\textbf{Type}\end{tabular}	&\begin{tabular}[x]{@{}l@{}}\textbf{Total}\end{tabular} &\begin{tabular}[x]{@{}l@{}}\textbf{Malicious}\end{tabular} &\begin{tabular}[x]{@{}l@{}}\textbf{C\&C}\end{tabular} &\begin{tabular}[x]{@{}l@{}}\textbf{Attack}\end{tabular} &\begin{tabular}[x]{@{}l@{}}\textbf{Normal}\end{tabular}\\ \hline
			10 & neris &\makecell[l]{IRC, Spam,\\CF} & 2,824,636 & 40,961 (1.45\%) & 341 (0.01\%) & 40,620 (1.44\%) &30,387 (1.07\%)\\ \hline
			11 & neris &\makecell[l]{IRC, Spam,\\CF} & 1,808,122 & 20,941 (1.16\%) & 673 (0.04\%) & 20,268 (1.12\%) &9,120 (0.5\%)\\ \hline
			12 & rbot &\makecell[l]{IRC, PS,\\US} & 4,710,638 & 26,822 (0.57\%) & 63 (0.00\%) & 26,759 (0.57\%) &116,887 (2.48\%)\\ \hline
			15 & rbot &\makecell[l]{IRC, DDoS,\\US} & 1,121,076 & 2,580 (0.23\%) & 52 (0.00\%) & 2,528 (0.23\%) &25,268 (2.25\%)\\ \hline
			15-2 & virut &\makecell[l]{Spam, PS,\\HTTP} & 129,832 & 901 (0.69\%) & 24 (0.02\%) & 877 (0.68\%) &4,679 (3.6\%)\\ \hline
			16 & menti &PS & 558,919 & 4,630 (0.83\%) & 199 (0.04\%) & 4,431 (0.79\%) &7,494 (1.34\%)\\ \hline
			16-2 & sogou &HTTP & 114,077 & 63 (0.06\%) & 26 (0.02\%) & 37 (0.03\%) &1,677 (1.47\%)\\ \hline
			16-3 & murlo &PS & 2,954,230 & 6,127 (0.21\%) & 1,074 (0.04\%) & 5,053 (0.17\%) &72,822 (2.46\%)\\ \hline
			17 & neris &\makecell[l]{IRC, Spam,\\CF, PS} & 2,087,508 & 184,987 (8.86\%) & 2,973 (0.14\%) & 182,014 (8.72\%) &43,340 (1.57\%)\\ \hline
			18 & rbot &\makecell[l]{IRC, DDoS,\\US} & 1,309,791 & 106,352 (8.12\%) & 33 (0.00\%) & 106,319 (8.12\%) &15,847 (1.2\%)\\ \hline
			18-2 & rbot &\makecell[l]{IRC, DDoS,\\US} & 107,251 & 8,164 (7.61\%) & 2 (0.00\%) & 8,162 (7.61\%) &2,718 (2.53\%)\\ \hline
			19 & nsys.ay &P2P & 325,471 & 2,168 (0.67\%) & 25 (0.01\%) & 2,143 (0.66\%) &7,628 (2.35\%)\\ \hline
			15-3 & virut &\makecell[l]{Spam, PS,\\HTTP} & 1,925,149 & 40,003 (2.08\%) & 536 (0.03\%) & 39,467 (2.05\%) &31,939 (1.65)\\ \hline
	\end{tabular}
\end{table}

The CTU-13 aims to reproduce the percentage of C\&C and attack flows according to real cases, where the number of malicious flows are very low in comparison to normal and background flows, resulting into a very imbalanced and challenging data set for botnet detection.

An Exploratory Data Analysis (EDA) reveals that distributions of original features of the CTU-13 are skewed and present an overlapping between normal and anomalous flows. Following we present the Figure \ref{fig:fig01}, that presents the distribution of flow duration for scenario 12, and exemplifies the data skewness and overlapping.

\begin{figure}[h!]
     \centering
     \includegraphics[width=10cm]{figures/raw_distplot_capture20110812_Dur.png}
     \caption{Distribution of flow duration for scenario 12.}
     \label{fig:fig01}
\end{figure}

>>>>

\chapter{Moment Distances from Robust Subspace for Network Anomaly Detection}
\label{ch:4_m_rpca}

@@

Some widely adopted algorithms for anomaly detection assume a gaussian distributed data, however this assumption may not be observed in some real world problems, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{lakhina2005mining,benson2010network}.

Findings of Benson \emph{et al.}  \cite{benson2010network} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. Leon-Garcia \cite{leon2017probability} also argues that Pareto distribution has been found to capture the behavior of many quantities of interest in the study of Internet behavior. Moreover, Benson \emph{et al.}  \cite{benson2010network} observes that the lognormal distribution is the best fit to model arrival processes in a data center.

This findings show that the skeweness and heavy-tailed distributions may be important for network traffic analysis, and can motivate evaluations to study the impact of skewed data into algorithms that rely on gaussian distributed data, as well as it can highlight opportunities to  exploit the skeweness and heavy-tailed distributions in order to obtain accurate classifiers for network anomaly detection.

Network anomaly detection problems are usually characterized by imbalanced data \cite{Phua2004minority,he2008learning,benson2010network}. However, learning algorithms for imbalanced data has been a challenging research topic, considering that the fundamental issue with the imbalanced learning problem is the ability of imbalanced data to significantly compromise the performance of most standard learning algorithms \cite{he2008learning}. Therefore, learning methods for imbalanced and skewed data have been attracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}.

We believe that the skewness of anomalous and normal traffic can highlight features for improving anomaly detection in imbalanced data, 
and that the distance between robust estimates of normal traffic and anomalies can highlight discrepancies and be used for network attack detection. Therefore, we propose the md-RPCA, which is an approach based on distances from moments computed from robust subspace learned by Robust Principal Component Analysis, in order to detect anomalies from network traffic. The proposed approach relies on a robust subspace of normal traffic for estimating higher moments. The anomaly detection from new observations evaluate the Mahalanobis distance between the robust higher moments and the higher moments of new observations, in a semi-supervised or unsupervised fashion, without the computational cost of new robust subspace learning for new observations.

We evaluate the accuracy of the md-RPCA for anomaly detection on synthetic data set and on the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}. The Experimental evaluation compares our proposal to widely adopted algorithms for anomaly detection based on clustering and statistical approaches.

% We also evaluate the results of ROBPCA \cite{hubert2005robpca}, which is a method that also relies on robust estimates with adjusted outlyingness based on robust skewness.

The main contribution of this work is a novel semi-supervised and unsupervised method for anomaly detection in skewed and imbalanced data, with results of experimental evaluation on synthetic and real data sets.

This Chapter is organized as follows. In Section \ref{sec:4_relatedworks}, it is conducted a literature review of network anomaly detection, botnet detection, and imbalanced learning. Section \ref{sec:4_datamodel} presents the data model and the evaluated data set. Section \ref{sec:4_proposal} describes the proposed approach for network attack detection. Section \ref{sec:4_experiments} discusses the experimental validation and Section \ref{sec:4_results} presents the results. Finally, Section \ref{sec:4_conclusion} draws the conclusions and the suggestions for future work.

@@

% TODO - re-write this, it is strange
% network anomaly detectiom
Bhuyan \emph{et al.} \cite{bhuyan2014network} provide an overview of facets of network anomaly detection, present attacks normally encountered by network intrusion detection systems, and categorize existing network anomaly detection methods and systems based on the underlying techniques. Ahmed \emph{et al.} \cite{ahmed2016survey} present an analysis of four major categories of anomaly detection techniques which include classification, statistical, information theory and clustering. Moustafa \emph{et al.} \cite{moustafa2019holistic} discuss aspects of anomaly-based Network Intrusion Detection Systems (NIDSs), explains cyber-attacks and new solutions for anomaly detection, and provides a benchmark data sets for training and validating approaches for detection.

@@

The objective of Garcia \emph{et al.} \cite{garcia2014empirical} is to compare three botnet detection methods using a simple and reproducible methodology, a good data set and a new error metric. This paper evaluates some data sets for network anomaly detection, surveys some approaches for botnet detection, proposes two methods (BClus and CAMNEP) for botnet detection and compare results to BotHunter \cite{gu2007bothunter}.

Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by attack, normal and background labeled data, in an imbalanced distribution like in a real network. The authors recommend scenarios for training and testing, in the way that none of the botnet families used in the training and cross-validation data set should be used in the testing data set, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Adopting the training and testing approach proposed by Garcia \emph{et al.}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

@@

% Anomaly detection by robust subspace learning
The problem of PCA or subspace learning for outlier corrupted data is called Robust Principal Component Analysis (RPCA) or robust subspace learning \cite{candes2011robust, vaswani2018robust}. RPCA aims to be resilient to outliers by means of a robust subspace learning \cite{vaswani2018robust} for outlier corrupted data, decomposing a given data matrix $\textbf{X}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, which refers to outliers’ matrix. This definition is also referred to as the sparse + low rank (S + LR) formulation \cite{vaswani2018robust}. We refer to \cite{lerman2018overview} and \cite{vaswani2018robust} for more details regarding robust subspace learning.

RPCA is been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adopted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of mean and covariance matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness.

Robust subspace learning has receiving a growing attention of researchers aiming the development of network anomaly detection systems \cite{rousseeuw1984mcd, rousseeuw1999fastmcd, hubert2005robpca,hubert2009robustskewed, pascoal2012robust, zhou2017anomaly}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}. Pascoal \emph{et al.} \cite{pascoal2012robust} proposed an approach based on a robust mutual information estimator for feature selection and based on RPCA for outlier detection in internet traffic. The anomaly detection proposed by Pascoal \emph{et al.} is an unsupervised approach that estimate the first $k$ robust principal components, calculate the score and the orthogonal distances, calculate the thresholds and classify new observations accordingly. Zhou and Paffenroth \cite{zhou2017anomaly} proposed the Robust Deep Autoencoders (RDA), which the central idea is that a RDA inherits the non-linear representation capabilities of autoencoders combined with the anomaly detection capabilities of RPCA. Considering that outliers and noise may reduce the quality of representations discovered by deep autoencoders, the proposed model isolates noise and outliers in the input by means of a RPCA approach, and the autoencoder is trained after this isolation. RDA was evaluated by the authors for the MNIST data set.

Benson \emph{et al.} \cite{benson2010network} conducted an empirical study of the network traffic in 10 data centers belonging to three different types of organizations, including university, enterprise, and cloud data centers. Findings of Benson \emph{et al.} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}.

We propose an approach based on higher moments distances from robust subspace for botnet detection. The proposed approach relies on a robust subspace of normal traffic for estimating higher moments. Thus, we propose that the anomaly detection for new observations should evaluate the Mahalanobis distance between the robust higher moments and the higher moments of new observations, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations, or in a unsupervised approach.

@@

This section also presents a description of the synthetic data model in Subsection \ref{sec:4_synthetic} and in Subsection \ref{sec:4_CTU-13} we describe the data model for the CTU-13 data set.

\subsection{The Synthetic data set}
\label{sec:4_synthetic}

In order to evaluate the null hypothesis $H_{0rob.sub}$, which assumes that the robust subspace learning does not improves the anomaly detection from imbalanced and skewed, we create two synthetic data sets characterized by skewed and heavy tailed distributions. We selected Pareto and Lognormal distributions to evaluate the anomaly detection capacity of our proposal, considering that these distributions have been adopted to represent network traffic in the internet and data centers \cite{benson2010network,leon2017probability}.

We adopt a contamination rate of 33\% to reproduce the imbalanced data of anomaly detection problems, and select the gaussian distribution for the addictive anomalies into Pareto and Lognormal distributions. Therefore, we generate $\boldsymbol{Y}_p$ and $\boldsymbol{Y}_l$ with 2,400 samples and two features, where 793 samples are gaussian anomalies, also referred as noise or contamination, while 1,607 samples are legitimate signals. The data set $\boldsymbol{Y}_p$ is composed by random Pareto distribution with random gaussian noise, while $\boldsymbol{Y}_l$ is composed by random Lognormal distribution with random gaussian noise.

The Figure \ref{fig:4.03} shows the histogram and the PDF of the Pareto distribution for two features of $\boldsymbol{Y}_p$.

\begin{figure}[!htb]
	\centering
	\subfigure[Feature 1]{%
		\includegraphics[width=0.48\textwidth]{figures/pareto1.png}
		\label{fig:4.01}
	}
	\subfigure[Feature 2]{%
		\includegraphics[width=0.48\textwidth]{figures/pareto2.png}
		\label{fig:4.02}
	}
	\caption[Pareto Distribution]{Pareto Distribution}
	\label{fig:4.03}
\end{figure}

The Figure \ref{fig:4.06} shows the histogram and the PDF of the Lognormal distribution for two features of $\boldsymbol{Y}_l$.

\begin{figure}[!htb]
	\centering
	\subfigure[Feature 1]{
		\includegraphics[width=0.48\textwidth]{figures/lognormal3.png}
		\label{fig:4.04}
	}
	\subfigure[Feature 2]{
		\includegraphics[width=0.48\textwidth]{figures/lognormal4.png}
		\label{fig:4.05}
	}
	\caption[Lognormal Distribution]{Lognormal Distribution}
	\label{fig:4.06}
\end{figure}

The Pareto distribution contaminated by gaussian noise is denoted as $\boldsymbol{Y}_p^c$, while $\boldsymbol{Y}_l^c$ denotes the Lognormal distribution contaminated by gaussian noise. Taking into account that semi-supervised algorithms requires training from normal data, we also generate $\boldsymbol{Y}_p^t$, while $\boldsymbol{Y}_l^t$, which are data sets of Pareto and Lognormal distributions for training, without contamination and composed by 1,607 samples and two features.

\subsection{The CTU-13 data set}
\label{sec:4_CTU-13}

The CTU-13 is a data set of botnet traffic that was captured in the Czech Technical University \cite{garcia2014empirical}, by means of a testbed and execution of malwares in a real network. The CTU-13 data set contains 13 scenarios with network flows of botnet malwares, that are: neris, rbot, virut, menti, sogou, nsys.ay and murlo. The botnet traffic is also classified as attack or command and control (C\&C), while the legitimate flows can also be classified as normal or background. It is important to note that the CTU-13 data set can also be formulated as a superposition of signal, artefact and noise, which refer to background, normal a botnet traffic, respectively.

The types of C\&C and attack flows present in CTU-13 data set are:

@@

Table \ref{tab:4.01} presents an overview organized by scenario and shows the malwares used for botnet attacks, the types of attacks and C\&C types, the total number of flows, the number of malicious flows, which includes flows of C\&C and attacks, and finally shows the number of normal flows.

\begin{table}[h!]
	\scriptsize
	\caption{CTU-13 data set Description}
	\label{tab:4.01}
	\begin{tabular}{| l | l | l | r | r | r | r | r | r | r | r | }
		\hline \rowcolor{Gray} \begin{tabular}[x]{@{}l@{}}\textbf{ID}\end{tabular}	&\begin{tabular}[x]{@{}l@{}}\textbf{Bot}\end{tabular}	 &\begin{tabular}[x]{@{}l@{}}\textbf{Type}\end{tabular}	&\begin{tabular}[x]{@{}l@{}}\textbf{Total}\end{tabular} &\begin{tabular}[x]{@{}l@{}}\textbf{Malicious}\end{tabular} &\begin{tabular}[x]{@{}l@{}}\textbf{C\&C}\end{tabular} &\begin{tabular}[x]{@{}l@{}}\textbf{Attack}\end{tabular} &\begin{tabular}[x]{@{}l@{}}\textbf{Normal}\end{tabular}\\ \hline
			10 & neris &\makecell[l]{IRC, Spam,\\CF} & 2,824,636 & 40,961 (1.45\%) & 341 (0.01\%) & 40,620 (1.44\%) &30,387 (1.07\%)\\ \hline
			11 & neris &\makecell[l]{IRC, Spam,\\CF} & 1,808,122 & 20,941 (1.16\%) & 673 (0.04\%) & 20,268 (1.12\%) &9,120 (0.5\%)\\ \hline
			12 & rbot &\makecell[l]{IRC, PS,\\US} & 4,710,638 & 26,822 (0.57\%) & 63 (0.00\%) & 26,759 (0.57\%) &116,887 (2.48\%)\\ \hline
			15 & rbot &\makecell[l]{IRC, DDoS,\\US} & 1,121,076 & 2,580 (0.23\%) & 52 (0.00\%) & 2,528 (0.23\%) &25,268 (2.25\%)\\ \hline
			15-2 & virut &\makecell[l]{Spam, PS,\\HTTP} & 129,832 & 901 (0.69\%) & 24 (0.02\%) & 877 (0.68\%) &4,679 (3.6\%)\\ \hline
			16 & menti &PS & 558,919 & 4,630 (0.83\%) & 199 (0.04\%) & 4,431 (0.79\%) &7,494 (1.34\%)\\ \hline
			16-2 & sogou &HTTP & 114,077 & 63 (0.06\%) & 26 (0.02\%) & 37 (0.03\%) &1,677 (1.47\%)\\ \hline
			16-3 & murlo &PS & 2,954,230 & 6,127 (0.21\%) & 1,074 (0.04\%) & 5,053 (0.17\%) &72,822 (2.46\%)\\ \hline
			17 & neris &\makecell[l]{IRC, Spam,\\CF, PS} & 2,087,508 & 184,987 (8.86\%) & 2,973 (0.14\%) & 182,014 (8.72\%) &43,340 (1.57\%)\\ \hline
			18 & rbot &\makecell[l]{IRC, DDoS,\\US} & 1,309,791 & 106,352 (8.12\%) & 33 (0.00\%) & 106,319 (8.12\%) &15,847 (1.2\%)\\ \hline
			18-2 & rbot &\makecell[l]{IRC, DDoS,\\US} & 107,251 & 8,164 (7.61\%) & 2 (0.00\%) & 8,162 (7.61\%) &2,718 (2.53\%)\\ \hline
			19 & nsys.ay &P2P & 325,471 & 2,168 (0.67\%) & 25 (0.01\%) & 2,143 (0.66\%) &7,628 (2.35\%)\\ \hline
			15-3 & virut &\makecell[l]{Spam, PS,\\HTTP} & 1,925,149 & 40,003 (2.08\%) & 536 (0.03\%) & 39,467 (2.05\%) &31,939 (1.65)\\ \hline
	\end{tabular}
\end{table}

The CTU-13 data set originally contains the following features:

\begin{itemize}
	\item Start Time;
	\item End Time;
	\item Duration;
	\item Source IP Address;
	\item Source Port;
	\item Direction;
	\item Destination IP Address;
	\item Destination Port;
	\item State;
	\item Type of service from source to destination;
	\item Type of service from destination to source;
	\item Total of Packets;
	\item Total of Bytes;
\end{itemize}

Our analysis of the available features leads to discard some features, considering that highly correlated features can bias or not improve the model, and that source or destination IP addresses can insert some false bias into learning models, since they can be changed by IP spoofing or other techniques.

TODO: talk about the feature distribution, that arent similar to pareto or lognormal, but it is not normal too
